---
title: "p8105_hw3_yc4584"
author: "Yingyu Cui"
date: "2024-10-08"
output: github_document
---

```{r setup and figure preferrences}
library(tidyverse)
library(patchwork)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# Problem 1
```{r load data}
library(p8105.datasets)
data("ny_noaa")
```


```{r NA data check}
na_counts = colSums(is.na(ny_noaa))

print(na_counts)
```

- A short discription:
There are 2595176 rows and 7 columns in this data frame. The data frame contains the information of the weather in New York from January 1, 1981 through December 31, 2010. The key variables are "date", "temperature_min", "temperature_max", "precipitation", "snowfall", "snow_depth". In all of the variables, the "prcp", "snow" and "snwd" variables are numeric variables and the "tmax", "tmin" and "id" are recognized as character variables. However, I would prefer to transfer "tman" and "tmin" into numeric characters. The missing data in every column is shown in the following table. 
     id    date    prcp    snow    snwd    tmax    tmin 
      0       0  145838  381221  591786 1134358 1134420 
We can see that there are around half of the data missing in the "tmax" and "tmin" columns. In "prcp", "snow" and "snwd" columns, there are less data missing, but still a large amount of data missing.

Question 1:
- Create separate variables for year, month, and day
```{r create variables}
ny_noaa = ny_noaa |> 
  mutate(year = year(date),
         month = month(date),
         day = day(date))
```

- Ensure observations for temperature, precipitation, and snowfall are given in reasonable units
we want the units for "prcp", "snow" and "snwd" to be tenths of mm, and the units for "tmax" and "tmin" to be in Celsius.  
```{r change units}
ny_noaa = ny_noaa |>
  mutate(tmax = as.numeric(tmax)) |>
  mutate(tmin = as.numeric(tmin)) |>
  mutate(
    prcp = ifelse(is.na(prcp), NA, prcp * 1/10),
    snow = ifelse(is.na(snow), NA, snow * 1/10),
    snwd = ifelse(is.na(snwd), NA, snwd * 1/10),
    tmax = ifelse(is.na(tmax), NA, (tmax - 32) * 5/9),
    tmin = ifelse(is.na(tmin), NA, (tmin - 32) * 5/9)
  )
```

-  most commonly observed values for snowfall
```{r most commonly observed values for snowfall}
common_snowfall = ny_noaa |> 
  count(snow) |> 
  arrange(desc(n))

print(common_snowfall)
```
We could find out that the most commonly observed value for snowfall is 0.0, which is 381221 times. I think this is because that snowfall is not a daily occurrence in most regions. Many places experience long periods without any snowfall, especially during non-winter months. Even in regions where snow is common, there are often days when no snow falls. 


Question2:
- Make a two-panel plot showing the average max temperature in January and in July in each station across years.
```{r two-panel plot showing the average max temperature}
ny_noaa_filtered_plot = ny_noaa |> 
  filter(month %in% c(1, 7)) |> 
  group_by(id, year, month) |> 
  summarize(avg_tmax = mean(tmax, na.rm = TRUE)) |> 
  ggplot(aes(x = year, y = avg_tmax)) +
  geom_point() +
  facet_grid(. ~ month) +
  labs(
    title = "Average Maximum Temperature in January and July",
    x = "Year",
    y = "Average Max Temperature (°C)"
  ) 
print(ny_noaa_filtered_plot)
```

- A short discription:Is there any observable / interpretable structure? Any outliers?
From the plot, we could see that the average maximum temperature in January is lower than that in July. The average maximum temperature in July is more stable than that In January. There are some outliers in the plot, but they are not very significant. The outliers in the January plot are approximately equally distributed in the lower and higher parts of the plot. The outliers in the July plot are mainly distributed in the lower part of the plot.


Question3:
- Part (i): tmax vs tmin plot using hexbin plot
```{r tmax vs tmin plot}
p1 = ggplot(ny_noaa, aes(x = tmin, y = tmax)) +
  geom_hex() +
  labs(
    title = "tmax vs tmin",
    x = "Minimum Temperature (°C)",
    y = "Maximum Temperature (°C)",
    fill = "Count"
  ) 
print(p1)
```
  Part (ii): Distribution of snowfall values > 0 and < 100, separated by year
```{r Distribution of snowfall values}
p2 = ny_noaa |> 
  filter(snow > 0 & snow < 100) |> 
  ggplot(aes(x = snow)) +
  geom_histogram(fill = "skyblue", color = "black") +
  facet_grid(. ~ year) +
  labs(
    title = "Distribution of Snowfall (0 < snowfall < 100)",
    x = "Snowfall (tenths of mm)",
    y = "Count"
  )
print(p2)
```

Combination of p1 and p2
```{r Combination of p1 and p2}
library(gridExtra)
grid.arrange(p1, p2, nrow = 1) 
```




